---
title: 'Peer Assessment #1'
author: "Lance Smith"
date: "Thursday, July 17, 2014"
output: html_document
---

 This is an analysis of the the step data collected over a period 
 of 61 days
 
 Function Block

 getStats - takes a dataframe and calculates the mean, median and sum of the steps column. It throws out any observation with zero steps. It returns a character vector with the three values
 
 convertToDF - takes a list of character vectors and converts them to a dataframe
```{r,echo=TRUE}
getStats <- function(x) {
        day <- x[1,2]
        steps <- length(x$steps)
        #zeros <- nrow(x[which(x$steps == 0),])
        intervalsWithSteps <- x[which(x$steps != 0),]
        meanSteps <- mean(intervalsWithSteps$steps)
        theMedian <- median(intervalsWithSteps$steps)
        theSum <- sum(intervalsWithSteps$steps)
        c(date=day,mean=meanSteps,median=theMedian,total=theSum)
}

# convert list into dataframe
convertToDF <- function(x) {
        theDay <- character()
        theMean <- numeric()
        theMedian <- numeric()
        theSum <- numeric()
        for (i in daytotals) {
                df <- i
                names(df) <- NULL
                theDay <- append(theDay,df[1])
                theMean <- append(theMean,as.numeric(df[2]))
                theMedian <- append(theMedian,as.numeric(df[3]))
                theSum <- append(theSum,as.numeric(df[4]))
        }
        return(data.frame(day=theDay,mean=round(theMean,2),median=round(theMedian,2),Total=theSum,stringsAsFactors=F))
}

```


<b>Part 1 - Loading and preprocessing the data.</b>
<br>Edit the first line below to set the working directy to the location of the data file

```{r,echo=TRUE}
setwd("~/Dropbox/Coursera/JHDS/Rresearch")
data <- read.csv("activity.csv",stringsAsFactors=F)

# Check the data
dim(data)
str(data)

```
 Preprocess the data. For the first part of the analysis, I am splitting the data into two groups: one with complete cases and one without
 
```{r,echo=TRUE}
good <- complete.cases(data)
goodData <- data[good,]
badData <- data[!good,]
```
 How many days of good data?
```{r,echo=TRUE}
completeDays <- length(unique(goodData$date))
completeDays
```
 <b>Part 2 - What is the mean total number of steps take per day?</b>
 
 Split the dataset by date

```{r,echo=TRUE}
days.list <- split(goodData,goodData$date)
```
 What is the mean total number of steps taken per day? The getStats function ignores zero days in the calculation. I did that to give meaningful numbers to the mean and median. Otherwise, they would have been close to or at zero for most days.
```{r,echo=TRUE}
daytotals <- lapply(days.list,getStats)
dfTotals <- convertToDF(daytotals)
```
Mean and median for each good day. 
```{r,echo=TRUE}
dfTotals
```
Histogram
```{r,echo=TRUE}
hist(dfTotals$Total,xlab="Interval",main="Steps per Interval (NAs removed) ",freq=T,breaks=20)
```
 <p>Part 3 What is the average daily activity
 
```{r,echo=TRUE}
avgSteps <- tapply(goodData$steps,goodData$interval,mean)
intervals <- as.integer(names(avgSteps))
plot(intervals,avgSteps,type="l")
```
<p>The inteval with the max number of steps is 835. The average number of steps take is 206 in this interval

```{r,echo=TRUE}
avgSteps[avgSteps == max(avgSteps)]
```
<b>Part 4 Imputing missing values</b>

Look at the good data
```{r,echo=TRUE}
dim(goodData)
str(goodData)
head(goodData)
```
Look at the bad data
```{r,echo=TRUE}
dim(badData)
str(badData)
head(badData)
```
All the bad data seems to be in the steps column. Which days have no data?
```{r,echo=TRUE}
unique(badData$date)
```
Verify these dates are not id the good data
```{r,echo=TRUE}
setdiff(badData$date,goodData$date)
```
Total number of missing values
```{r,echo=TRUE}
missing <- nrow(badData)
```
Impute stratgy - use the 5 minute average and adjust the value using div/mod to find the floor/ceiling
```{r,echo=TRUE}
imputed <- sapply(avgSteps,function(x) {
        div <- x%/% 1
        mod <- x%% 1
        if(mod > 0.5) {
                div <- div + 1
        }
        return(div)
})
```
Add the imputed data to the badData frame. Since there are 8 days of bad data, just replicate the imputed data 8 times and apply it to the dataframe
```{r,echo=TRUE}
badData$steps <- rep(imputed,8)
head(badData,10)
```
Create a new dataset with no missing values
```{r,echo=TRUE}
imputedData <- rbind(goodData,badData)
```
Get the median,mean and histogram
```{r,echo=TRUE}
days.list <- split(imputedData,imputedData$date)
daytotals <- lapply(days.list,getStats)
imputedTotals <- convertToDF(daytotals)
```
Mean,median and histogram
```{r,echo=TRUE}
imputedTotals
```
 Adding the imputed totals back into the dataset has distorted the graph. Intervals with lower averages have been depressed while those with higher average have been inflated. 
 
```{r,echo=TRUE}
hist(imputedTotals$Total,xlab="Interval",main="Steps per Interval (NAs imputed) ",freq=T,breaks=20)
```
<p>

<b>Part 5 Are there differences in activity patterns between weekdays and weekends?</b>
<br>
Move date string to data objects and create a "day" factor using the weekdays function
```{r,echo=TRUE}
days <- as.Date(imputedData$date)
imputedData$date <- days
imputedData$day <- as.factor(weekdays(imputedData$date))
```
Split the data set into weekdays and weekends
```{r,echo=TRUE}
weekends <- imputedData[imputedData$day %in% c("Saturday","Sunday"),]
weekdays <- imputedData[!imputedData$day %in% c("Saturday","Sunday"),]
```

Create a weekend averages dataframe
```{r,echo=TRUE}
avgStepsWeekend <- tapply(weekends$steps,weekends$interval,mean)
SatSun <- as.integer(names(avgSteps))
intervalData1 <- data.frame("average_steps"=avgStepsWeekend,"interval"=SatSun,"day_type"="Weekend")
```

Create a weekday averages dataframe
```{r,echo=TRUE}
avgStepsWeekday <- tapply(weekdays$steps,weekdays$interval,mean)
MonFri <- as.integer(names(avgSteps))
intervalData2 <- data.frame("average_steps"=avgStepsWeekday,"interval"=MonFri,"day_type"="Weekday")
```
Bind them together
```{r,echo=TRUE}
intervalDataAll <- rbind(intervalData1,intervalData2)
```
Create a two level factor from the "weekend,weekday" character strings
```{r,echo=TRUE}
intervalDataAll$day_type <- as.factor(intervalDataAll$day_type)
```

Use the new factor to split the graph into two facets
```{r,echo=TRUE}
library(ggplot2)
 qplot(interval,average_steps,data=intervalDataAll,facets=day_type~.,geom=c("line"),
      xlab="Intervals",ylab="Number of Steps")
```
<p>Based on the graphs, there is more activity during the afternoon hours on the weekends. While the highest activity is a spike during the
weekday, the weekend sees more consistant activity durning the day.

